---
output: github_document
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
set.seed(123)
```

# Install

Install the development version from GitHub:

```{r, eval = FALSE}
remotes::install_github("giuseppec/customtrees")
```

# Objectives
```{r}
library(devtools)
library(tidyverse)
library(Rmalschains)
library(iml)
library(ranger)
library(kmlShape)
library(dtw)
load_all()
```

```{r}
# objective that fits a constant in the nodes (CART) 
SS = function(x, y) {
  ypred = mean(y)
  sum((y - ypred)^2)
}

# objective that fits a linear model in the nodes (mob)
SS_lm = function(x, y) {
  ypred = predict(lm(y ~ x))
  sum((y - ypred)^2)
}

# objective for multivariate targets (multivariate tree), see MultivariateRandomForest::Node_cost function
SS_mah = function(x, y, cov) {
  center = colMeans(y)
  # cov = cov(y) we need to pass the cov of all data
  sum(mahalanobis(y, center = center, cov = cov, tol = 1e-30))
}

# Frechet distance FDA measure
SS_fre = function(x, y) { # slow
  # using only y-axis of curves is enough as x-axis is always the same for all curves
  require(kmlShape)
  center = colMeans(y)
  grid.x = as.numeric(names(center))
  pdp.y = unname(center)
  dist = apply(y, 1, function(ice) distFrechet(grid.x, pdp.y, grid.x, ice, FrechetSumOrMax = "sum"))
  sum(dist)
}

# Dynamic time warping FDA measure
SS_dtw = function(x, y) {
  require(dtw)
  pdp = colMeans(y) # this is the pdp
  dist = apply(y, 1, function(ice) dtw(ice, pdp, distance.only = TRUE)$normalizedDistance)
  sum(dist)
}
```

# Notes

- This package is not intended to be fast. It serves as a modular framework and playground to explore/study the splitting of features by custom objectives.
- Splits for categorical variables currently not implemented and tested. Try to handle categoricals as numerics as workaround. 
- The `perform_split` function computes (and aggregates) the objective in the generated nodes after splitting w.r.t. specific split points.
- Binary splits generate two nodes and are implemented in `find_best_binary_split`. The implementation  does exhaustive search of split point candidates to find the best split point for a given feature. 
- Multiple splits generate multiple nodes and are implemented in `find_best_multiway_split`. The implementation currently uses a slow simulated annealing optimization to find the best split point for a given feature (might be improved and replaced with other, faster optimization procedures).


# CART with binary splits (constant model in node)

```{r}
nsim = 1000L
x = x = sort(runif(n = nsim, min = 0, max = 2*pi))
q = quantile(x, seq(0, 1, length.out = 100), type = 1)
y = ifelse(x > pi/2, rnorm(nsim, mean = 0), rnorm(nsim, mean = 10, sd = 2))
X = data.frame(x = x)

split = split_parent_node(y, X, objective = SS, optimizer = find_best_binary_split)
split

# plot result
plot(x, y)
abline(v = unlist(split$split.points))
```

# CART with multiple splits (constant model in node)

```{r}
y = ifelse(x < pi/2, rnorm(nsim, mean = 0), 
  ifelse(x < pi, rnorm(nsim, mean = 10, sd = 2), 
    rnorm(nsim, mean = -10, sd = 5)))

# Simulated annealing from find_best_multiway_split
split = split_parent_node(y, X, objective = SS, optimizer = find_best_multiway_split, 
  n.splits = 2, control = list(maxit = 1000))

# MA-LS Chains optimization from find_best_multiway_split2 (claimed to be faster and better)
split2 = split_parent_node(y, X, objective = SS, optimizer = find_best_multiway_split2, 
  n.splits = 2, control = malschains.control(istep = 100, ls = "sw"))

split
split2

plot(x, y)
abline(v = unlist(split$split.points))
```

# MOB with binary splits (linear model in node)

```{r}
y = 4 + 2 * cos(x) + rnorm(nsim, mean = 0, sd = abs(cos(x)) / 2)

split = split_parent_node(y, X, objective = SS_lm, optimizer = find_best_binary_split, n.splits = 1)
split

plot(x, y)
abline(v = unlist(split$split.points))
```

# MOB with multiple splits (linear model in node)

```{r}
y = 4 + 2 * cos(x*2) + rnorm(nsim, mean = 0, sd = abs(cos(x)) / 2)

split = split_parent_node(y, X, objective = SS_lm, optimizer = find_best_multiway_split2, 
  n.splits = 3, control = malschains.control(istep = 100, ls = "sw"))
split

plot(x, y)
abline(v = unlist(split$split.points))
```

# Group ICE Curves with Multivariate Tree (binary splits, constant model in node)

We first generate some functional data:

```{r}
# Simulate Data
n = 500
x1 = runif(n, -1, 1)
x2 = runif(n, -1, 1)
x3 = sample(c(0, 1), size = n, replace = TRUE, prob = c(0.5, 0.5))
x4 = sample(c(0, 1), size = n, replace = TRUE, prob = c(0.7, 0.3))
eps = rnorm(n, 0, 1)

# noisy vars
x5 = sample(c(0, 1), size = n, replace = TRUE, prob = c(0.5, 0.5))
x6 = rnorm(n, mean = 1, sd = 5)

y = 0.2*x1 - 8*x2 + ifelse(x3 == 0, I(16*x2),0) + ifelse(x1 > mean(x1), I(8*x2),0) + eps
# We also get interesting results using a 2-way interaction of numeric features
# y = 0.2*x1 - 8*x2 + 8*x6*x2 + eps

dat = data.frame(x1, x2, x3, x4, x5, x6, y)
X = dat[, setdiff(colnames(dat), "y")]

# Fit model and compute ICE for x2
mod = ranger(y ~ ., data = dat, num.trees = 500)
pred = predict.function = function(model, newdata) predict(model, newdata)$predictions
model = Predictor$new(mod, data = X, y = dat$y, predict.function = pred)
effect = FeatureEffects$new(model, method = "ice", grid.size = 20, features = "x2")

# Plot ICE curves: WE WANT TO FIND SUBGROUPS SUCH THAT ICE KURVES ARE HOMOGENOUS
ggplot(effect$results$x2, aes(x = .borders, y = .value)) + 
  geom_line(aes(group = .id))
```

Formulate curves above by multivariate target and find feature that splits the curves such that they are more homogenous in the nodes:

```{r}
# Get ICE values and arrange them in a horizontal matrix
Y = spread(effect$results$x2, .borders, .value)
Y = Y[, setdiff(colnames(Y), c(".type", ".id", ".feature"))]
str(X) # contains our feature values
str(Y) # contains ICE values for each grid point

# compute covariance for data and use this in for mahalanobis distance in the objective
COV = cov(Y)
SS_mah2 = function(x, y) SS_mah(x, y, cov = COV)
sp = split_parent_node(Y = Y, X = X, objective = SS_mah2, 
  n.splits = 1, optimizer = find_best_binary_split)
sp
node_index = generate_node_index(Y, X, result = sp)
str(node_index)

# frechet distance yields same splits but is a bit slower
sp_frechet = split_parent_node(Y = Y, X = X, objective = SS_fre,
  n.splits = 1, optimizer = find_best_binary_split)
sp_frechet
node_index_frechet = generate_node_index(Y, X, result = sp_frechet)
str(node_index_frechet)

## dynamic time warping distance yields same splits but is much slower
# sp_dtw = split_parent_node(Y = Y, X = X, objective = SS_dtw, 
#   n.splits = 1, optimizer = find_best_binary_split)
# sp_dtw
# node_index_dtw = generate_node_index(Y, X, result = sp_dtw)
# str(node_index_dtw)
```

```{r}
# Compare with MultivariateRandomForest yields same result
library(MultivariateRandomForest)
invcov = solve(cov(Y), tol = 1e-30)
sp2 = splitt2(X = as.matrix(X), Y = as.matrix(Y), m_feature = ncol(X), 
  Index = 1:nrow(X), Inv_Cov_Y = invcov, Command = 2, ff = 1:ncol(X))
str(sp2)
```

Visualize the results:

```{r}
plot.data = effect$results$x2
plot.data$.split = node_index$class[plot.data$.id]

ggplot(plot.data, aes(x = .borders, y = .value)) + 
  geom_line(aes(group = .id)) + facet_grid(~ .split)
```

# Group ICE Curves with Multivariate Tree (multiway splits, constant model in node)

Multiway split **fails** with `SS_mah2` (mahalanobis distance) as objective. 
This is because the curve structure along the x-axis is not considered in the distance calculation!

```{r}
sp_multiway = split_parent_node(Y = Y, X = X, objective = SS_mah2, 
  n.splits = 2, optimizer = find_best_multiway_split2)
sp_multiway
node_index_multiway = generate_node_index(Y, X, result = sp_multiway)
str(node_index_multiway)

plot.data$.split = node_index_multiway$class[plot.data$.id]

ggplot(plot.data, aes(x = .borders, y = .value)) + 
  geom_line(aes(group = .id)) + facet_grid(~ .split)
```

Instead, using a distance measure that is suited for curves (e.g., frechet distance) works:

```{r}
sp_multiway_frechet = split_parent_node(Y = Y, X = X, objective = SS_fre, 
  n.splits = 2, optimizer = find_best_multiway_split2)
sp_multiway_frechet
node_index_multiway_frechet = generate_node_index(Y, X, result = sp_multiway_frechet)
str(node_index_multiway_frechet)

plot.data$.split = node_index_multiway_frechet$class[plot.data$.id]

ggplot(plot.data, aes(x = .borders, y = .value)) + 
  geom_line(aes(group = .id)) + facet_grid(~ .split)
```

